---
title: "R Vector Analysis Lab"
author: "Bill Drummond"
date: "11/22/2021"
output: html_document
always_allow_html: true
---


# Setup the lab

Open R studio, open, and run vector-analysis-lab-setup.R to setup the lab, load the spatial datasets and run re-projections.

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warnings = FALSE)
library(tidyverse)
library(readxl)
library(sf)
library(tmap)

# setwd("c:/temp/class2421")

source("./functions/functions1c.R")

# read spatial datasets and re-project to wgs84 if needed
states = st_read("./gisdata/cb_2018_us_state_20m.shp") %>% 
  tolow() %>% 
  st_transform(4326) %>% 
  filter(!(stusps %in% c("AK", "HI", "PR")))

counties = st_read("./gisdata/cb_2018_us_county_20m.shp") %>% 
  tolow() %>% 
  st_transform(4326)
nercs = st_read("./gisdata/NercRegions_201907.shp") %>% 
  tolow() 
tlines = st_read("./gisdata/HVTranLines.shp") %>% 
  tolow() 
plants = st_read("./gisdata/PowerPlants_US_202004.shp") %>% 
  tolow() %>% 
  filter(total_mw >= 100) %>% 
  filter(!(statename %in% c("Alaska", "Hawaii", "Puerto Rico")))

# define sf projection with OGC WKT human-readable text strings
#   note the OGC WKT keywords shown in capitals

usalberswkt = 
'PROJCS["USA_Contiguous_Albers_Equal_Area_Conic_USGS_version",
    GEOGCS["GCS_North_American_1983",
        DATUM["D_North_American_1983",
            SPHEROID["GRS_1980",6378137.0,298.257222101]],
        PRIMEM["Greenwich",0.0],
        UNIT["Degree",0.0174532925199433]],
    PROJECTION["Albers"],
    PARAMETER["False_Easting",0.0],
    PARAMETER["False_Northing",0.0],
    PARAMETER["Central_Meridian",-96.0],
    PARAMETER["Standard_Parallel_1",29.5],
    PARAMETER["Standard_Parallel_2",45.5],
    PARAMETER["Latitude_Of_Origin",23.0],
    UNIT["Meter",1.0]]'

statesaea = states %>% st_transform(usalberswkt)

# defining a projection with a proj4 string
#    note the + as the lead character for each parameter

utm16 = "+proj=utm +zone=16 +ellps=GRS80 +datum=NAD83 +units=m +no_defs" 
statesutm16 = states %>% st_transform(utm16)

counties2 = counties %>% st_transform(usalberswkt)
nercs2 = nercs %>% st_transform(usalberswkt)
states2 = states %>% st_transform(usalberswkt)
tlines2 = tlines %>% st_transform(usalberswkt)
plants2 = plants %>% st_transform(usalberswkt)

```


-----

# Task 1: Spatial joins

An ordinary join matches records from two dataframes using columns of shared values. A spatial join matches features from two spatial dataframes using a geographic relationship between features; and it copies attributes from the joined spatial dataframe to the source spatial dataframe.

Assume we want to simply assign each US county to a NERC region.  We could conduct a long, complex polygon overlay, but an alternate quick method is to convert county polygons to centroids, then conduct a spatial join to join NERC attributes to each centroid.


```{r}

coucents2 = st_centroid(counties2)

counerc2 = coucents2 %>% st_join(nercs2, join = st_intersects)
counerc3 = st_drop_geometry(counerc2) %>% 
  select(geoid,nerc,nerc_label)
counerc4 = counties %>% left_join(counerc3)


```

1. What kind of features does each step in the analysis produce?
2. Why do we need to drop the geometry to create counerc3?
3. How would you alter this code to produce a single counerc output polygon dataset without creating counerc1, counerc2, and counerc4?

-----

We'll now plot the results, and plot a zoomed-in version for Texas.  Use CTRL-C and CTRL-V to paste this text into your R script, and run each piece of it.

Run each command below

```{r}

zoomx = function(sfin){
 box = st_bbox(sfin)
 xlimits = c(box$xmin,box$xmax)
 return(xlimits) }
 
zoomy = function(sfin){
 box = st_bbox(sfin)
 ylimits = c(box$ymin,box$ymax)
 return(ylimits) }

ggplot() +
  geom_sf(data=states2) +
  geom_sf(data=counerc4, aes(fill=nerc))

targstates = states2 %>% filter(name %in% c("Texas"))

ggplot() +
  geom_sf(data=states2) +
  geom_sf(data=counerc4, aes(fill=nerc)) +
  geom_sf(data=states2,fill=NA,size=2) +
  geom_sf(data=nercs2,fill=NA,color="purple") + 
  coord_sf(xlim=zoomx(targstates),ylim=zoomy(targstates))

```

Zooming to a map area is somewhat awkward in ggplot.  You use an extra + and command of the form:  
   
  coord_sf(xlim = c(-20, 45), ylim = c(30, 73)
  
However, for any sf dataset you can extract the bounding box with a command of the form

   box = st_bbox(states2)  
   
which produces output of the form

           xmin       ymin       xmax       ymax   
     -2356113.7   268660.9  2258154.4  3165721.7   
      
1. Examine the zoomx and zoomy functions above.  What do they accomplish?
2. Alter the code above so you zoom to Texas and three of its neighbors: New Mexico, Louisiana, and Oklahoma, and create that map.

-----

The spatial join operation has a number of different geographic relationships that can be used to join features.  If you use st_join without any special arguments, the actual command issued includes the default geographic relationship, which is st_intersects.  The command shown below is the one that is really executed when you type the counerc1 command in the sequence above.

counerc2 = coucents2 %>% st_join(nercs2, join = st_intersects)

You can substitute other geographic relationships, such as:

st_contains_properly  
st_contains  

st_covered_by  
st_covers 

st_crosses  

st_disjoint  

st_equals_exact  
st_equals  

st_is_within_distance  
st_nearest_feature  

st_overlaps 
st_touches  
st_within  

As discussed below in Task 6, st_intersects returns a set of TRUE/FALSE values as st_join tries to match each record in the first(input) dataframe to each record in the join (second) dataframe, using the specified (or default st_intersects) geographic relationship.

If the match attempt returns TRUE, an output record is created combining the attributes of the first dataframe and the attributes of the second dataframe.

-----

# Task 2: Buffers

Our next task is to identify areas within 10 miles (16.09 km) of a Georgia high voltage transmission line.

We will first clip the national tlines dataset to the Georgia boundary, project the tlines data into a projected coordinate system, buffer the lines, and re-project back to WGS84.

```{r}

georgia2 = states2 %>% filter(name == "Georgia")

gatlines2 = tlines2 %>% st_intersection(georgia2) 

ggplot() +
  geom_sf(data=georgia2) +
  geom_sf(data=gatlines2,color="orange")

```

Check the ggplot to confirm that the intersection/clip has worked correctly.

-----

We now create the buffer:

```{r}

gatbuffer2 = gatlines2 %>% st_buffer(dist=16093.4)

ggplot() +
  geom_sf(data=georgia2) +
  geom_sf(data=gatbuffer2,color="green")

gatbuffer3 = st_union(gatbuffer2)

ggplot() +
  geom_sf(data=georgia2) +
  geom_sf(data=gatbuffer3,color="green")

gatbuffer4 = st_sf(gatbuffer3) %>% 
  mutate(buffdist = "16093.4 meters")

ggplot() +
  geom_sf(data=georgia2) +
  geom_sf(data=gatbuffer4,color="green")


```

1. Compare the first, second, and third plots.  What does the st_union() command accomplish?
2. The second and third plots look identical.  But what are the object types of gatbuffer3 and gatbuffer4? 
3. What does the st_sf() command accomplish?  
4. Note: sf spatial analysis commands don't always return sf objects, even when that's what we would expect.  We need to pay attention to sf command output types.

-----

# Task 3: Point in polygon

We would now like to test whether most of Georgia's large power plants fall inside the 10-mile buffer around major transmission lines.

We first pull Georgia's plants from the national plant dataset.

The crucial operation is st_intersets(), which returns a matrix of TRUE or FALSE values depending on whether each object intersects another set of objects.  The sparse=FALSE clause ensures that both TRUE and FALSE values are returned, rather than only TRUE values.


```{r}
gaplants2 = plants2 %>% st_intersection(georgia2)

neartlines2 = gaplants2 %>% st_intersects(gatbuffer4, sparse=FALSE)
neartlines3 = data.frame(neartlines2)

gaplants3 = gaplants2 %>% bind_cols(neartlines3)

ggplot() +
  geom_sf(data=georgia2) +
  geom_sf(data=gatbuffer4,color="green") +
  geom_sf(data=gatlines2,color="orange") +
  geom_sf(data=gaplants3,aes(color=neartlines2))

```

1. Compare st_intersection() with st_intersects().  What type of data does each return?
2. Are most of the plants within 10 miles of major transmission lines?
3. What area of the state has a concentration of major lines, but very few plants?


-----

# Task 4: Dissolving (merging) polygons with common attribute values.

Our next spatial task: merging polygons that share an attribute value.  We wish to generate a polygon dataset with one polygon for each of the nine Census divisions, each of which consists of a number of states.  Georgia is in the South Atlantic division, which stretches from Georgia north to Maryland.

We are able to conduct the merge without using a special st command.  Instead we can use the dplyr group_by and summarize commands.  In one step we are able to both merge polygons and aggregate individual state numbers into division numbers.

We first read a dataset of state information.  It includes a record for each state and several fields, include the states division code and division name.

```{r}
stateinfo = read_excel("./data/stateinfo5.xls") %>% 
  rename(name = statename)

states3 = states2 %>% left_join(stateinfo)

divisions2 = states3 %>% group_by(divname) %>% 
  summarize(aland = sum(aland),
            awater = sum(awater))

ggplot() +
  geom_sf(data=divisions2,aes(fill=divname)) 

ggplot() +
  geom_sf(data=divisions2,aes(fill=divname)) +
  geom_sf(data=states2,alpha=0) 

```

1. Check the ggplot output for correctness.  In the second plot, what does alpha=0 do? Try alpha values of 0.5 and 1.0.
2. Why rename the statename variable to name during the creation of states2?
3. How have these commands handled aland and awater values?

-----

# Task 5: Polygon overlay to generate small polygons

Our final spatial analysis task is classic polygon overlay.  We will be overlaying state polygons and NERC polygons to generate small polygons that are "nested" within both states and NERCs.  This will allow us, for example, to calculate the area of each state within each NERC.

We first simplify the attributes for each input layer, then invoke st_intersection() to generate the small polygon dataset.

Finally we plot the results with ggplot and with a tmap interactive map.

```{r}

states4 = states2 %>% 
    select(stname=name, geoid)

nercs4 = nercs2 %>% 
  select(nercname = nerc)

overlay4 = nercs4 %>% 
  st_intersection(states4)

ggplot(data=overlay4) + 
  geom_sf(aes(fill=nercname))

tmap_mode("view")

tm_shape(overlay4) +
  tm_polygons(col="nercname",
  popup.vars = c("nercname","stname"))

```




