---
title: "egrid-lab-1a"
author: "Bill Drummond"
date: '2022-12-29'
output: html_document
---

The source command loads custom functions, which include:      
1. tolow(): converts all dataframe column names to legal, lower-case names  
2. mid():   extracts a substring from a string based on start position and length  
3. clip():  copies a dataframe into an enlarged copy buffer; like Ctrl-C


```{r, results=FALSE, message=FALSE, warning=FALSE, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warnings = FALSE)
library(tidyverse)
library(readxl)
library(sf)
library(tmap)

source("functions.R")


```


---




  
## Background on electricity, dataframes, and dplyr

### Electricity, power, and energy

- Power is how much energy is delivered at an individual moment in time.  It is similar to speed.
- Energy is how much power is delivered over time.  To simplify: power over time equals energy. It is similar to distance (speed over time equals distance)
- We measure electricity power in Watts: a 1,000 Watt microwave oven delivers a kilowatt (= 1,000 Watts) of power.
- We measure of electricity energy in kWh, or kilowatt hours.  1 kWh of energy is delivered by a 1,000 Watt microwave running for an hour.
- In the U.S. we can also measure energy in British thermal units or Btus.  One kWh of energy equals 3412.14 Btus.
- We describe electricity power plant size with MW (megawatts) of power.
- A very large coal or natural gas plant may be 3,000 MW (or 3 GW).
- A large solar farm in Georgia may be 200 MW.


### Dataframes and dplyr

- Dataframes are the way R stores data structured into rectangular arrays of rows and columns  
  - Rows are entities; columns are characteristics of the entities 

- dplyr is the main tidyverse package for manipulating dataframes  
  - The **%>%** command pipes data from one dplyr command to the next  
  - The **= (or <-)** command assigns a value to an object; in dplyr it sets the name of the final output of a series of dplyr commands  
  - The **filter()** dplyr command removes rows from the dataframe that do not meet a condition  
  - The **select()** dplyr command chooses columns from the dataframe and removes ones not chosen  
  - The **mutate()** dplyr command calculates one or more new variables  
  - The **ifelse()** function, used within mutate, defines a condition, the output value when the condition is true, and the output value when the condition is false

---

## Reads a spreadsheet to create egrid1

### read_excel

read_excel() can specify  
-   a specific sheet within Excel file, and 
-   the number of rows to skip before reading data, and 
-   assumes the first row has column names followed by rows of data  

```{r}
egrid1 = read_excel("egrid2020_data_metric.xlsx",
                    sheet="PLNT20",skip=1)
```


---

## Creates egrid2

- In a condition,  
  - R specifies "or" with |, 
  - specifies "and" with &,  
  - and specifies "not" with !  

- Note: giga is billions (10^9) mega is millions (10^6)
- Note: MW are units of power, MWh are units of energy
- Net generation (annual) is in MWh, and is converted to GWh by dividing by 10^6
- co2 equivalent emissions is in metric tons, and is converted to millions of metric tons by dividing by 10^6.
- Nameplate capacity is in MW, stays in MW.

```{r}

egrid2 = egrid1 %>% 
  tolow() %>% 
  filter(pstatabb == "GA" | pstatabb == "AL") %>% 
  mutate(heatinbtum = plhtian,
         netgengwh = plngenan/10^3,
         epaemimmt = plco2eqa/10^6) %>% 
  select(plantname = pname,
         stateabbr = pstatabb,
         fuel = plfuelct,
         nameplatecapmw = namepcap,
         heatinbtum,
         netgengwh,
         epaemimmt,
         lat, lon)
```


---

## Creates egrid3

- The ifelse() function takes inputs of  
  - T/F condition,  
  - value to return if condition is true, and  
  - value to return if condition is false  
- The is.na() function tests to see if a value is NA (Not Available, or missing)  
- power over time equals energy, so 1 MW of power delivered over an hour produces MWh of energy  
- capacity factor is the percentage of    
  - actual energy produced, divided by  
  - maximum possible energy produced according to nameplate capacity  
- heatinbtum is heat input (in btu millions) to generate electricity, which is the amount of energy released when burning fossil fuels or creating nuclear fission


```{r}
egrid3 = egrid2 %>% 
  mutate(nameplatecapmw = ifelse(is.na(nameplatecapmw), 0, nameplatecapmw),
         heatinbtum  =    ifelse(is.na(heatinbtum), 0, heatinbtum),
         netgengwh =      ifelse(is.na(netgengwh), 0, netgengwh),
         epaemimmt =      ifelse(is.na(epaemimmt), 0, epaemimmt),
         capfactor = 100 * ((netgengwh*10^3) / 
                            (nameplatecapmw*365*24)))
```

---

## Reads CO2 coefficients for egrid fuels  

- coefficients represent 
  - grams of CO2 released per 1,000 btu of fuel burned or
  - mmt of CO2 released per quadrillion btu of buel burned

```{r}
coef = read_excel("egrid-coef.xlsx")
```


---

# Creates egrid4

- This operation introduces left_join() to join columns from a second table based on matching values of a shared column or columns  
- R automatically searches for any shared column(s), finds the "fuel" columns, matches identical fuel values, and copies the matching data from the second dataframe to the first  
- Note 1: since a coefficient is grams of CO2 / 1000 btu, we convert millions of btu (btum) to thousands of btu when we divide by 10^3  
- Note 2: the result of multiplying the coefficient times btu thousands is grams of CO2.  There are 1 million (10^6) grams in a metric ton and one million metric tons in 1 mmt so we divide grams by 10^12 to calculate millions of metric tons of CO2 released
- Note 3: We also calculate the ratio of our calculated mmt to the EPA's mmt from the egrid spreadsheet

```{r}
egrid4 = egrid3 %>% 
  left_join(coef) %>% 
  mutate(calcemimmt = heatinbtum*10^3 * coef / 10^12,
         calctoepa = 100*round(calcemimmt/epaemimmt,4))
```

---

## Creates egrid4state by aggregating plant-level data to generate state-level data

- These operations introduce group_by() and summarize()  
  - group_by() divides the dataframe into separate subsets and can use one or more variables to create the subsets  
  - summarize() creates one output row/record for each subset and conducts whatever operation is specified (sum, mean, etc.) to "summarize" all the records in each subgroup  
  
```{r}
egrid4state = egrid4 %>% 
  group_by(stateabbr) %>% 
  summarize(netgengwh = sum(netgengwh,na.rm=T),
            epaemimmt = sum(epaemimmt,na.rm=T),
            calcemimmt = sum(calcemimmt,na.rm=T))

egrid4statefuel = egrid4 %>% 
  group_by(stateabbr,fuel) %>% 
  summarize(netgengwh = sum(netgengwh,na.rm=T),
            epaemimmt = sum(epaemimmt,na.rm=T),
            calcemimmt = sum(calcemimmt,na.rm=T))
```
  
## Creates egrid5 

- We use the lon and lat fields to build a spatial data frame of points
- The coordinate reference system is the ESPG number for common lat/lon coordinates


```{r}
egrid5 = egrid4 %>% 
         filter(!(is.na(lat) | is.na(lon))) %>% 
         st_as_sf(coords = c("lon", "lat"), 
                    crs = 4326) 
```


## Creating interactive maps with tmap package

- creates interactive map  
- alternate mode is "plot" for static map  


```{r}
tmap_mode("view")
```


- plots spatial dataframe points on Open Street Map basemap  

```{r}
tm_basemap("OpenStreetMap") +
  tm_shape(egrid5) +
  tm_dots()
```

- colors dots by fuel field and   
  - sizes dots by net generation in GWh
  
```{r}
tm_basemap("OpenStreetMap") +
  tm_shape(egrid5) +
  tm_dots(col="fuel",size="netgengwh")
```

- adds imagery basemap and  
  - sets layer opacity (opposite of transparency) to 50%
  
```{r}
tm_basemap(providers$Esri.WorldImagery) +
  tm_shape(egrid5) +
  tm_dots(col="fuel",size="netgengwh",alpha=.50)
```


- nests dplyr filter command within tm_shape()
  - to limit plants to solar ones, and
  - uses OpenStreetMap basemap tiles

```{r}
tm_basemap("OpenStreetMap") +
  tm_shape(egrid5 %>% filter(fuel=="SOLAR")) +
  tm_dots(col="yellow", size="nameplatecapmw")
```

- maps dots with size based on emissions and
- saves to a tmap object named m,
- then saves the tmap object to an interactive html file 8" high

```{r}
m = tm_basemap("OpenStreetMap") +
    tm_shape(egrid5) +
    tm_dots(col="fuel", size="calcemimmt")


tmap_save(m,"elecemissions.html",height=8)
```


